Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-07-26 manuscript/usses_water_ms.tex
addresses PBA minor comments

\usepackage{lipsum} \usepackage{setspace} \usepackage{todonotes}
\usepackage{rotating} \usepackage{color, soul} \usepackage{sectsty}
\usepackage{bm,mathrsfs}

@@

% 
\begin{abstract}
Precipitation is predicted to become more variable in the western U.S.,
meaning years of above and below average precipitation will become more
common. Such periods of drought and deluge are major drivers of plant
community dynamics and ecosystem functioning in water limited
grasslands. Here we report the results of a five-year experiment where
we used drought and irrigation treatments (50\% decrease/increase) to
see how a sagebrush steppe plant community in Idaho will respond to
future climate changes. Ecosystem functional response, defined as the
relationship between soil moisture and aboveground net primary
productivity, was suprisingly consistent across drought, control, and
irrigation treatments. Aboveground net primary productivity (ANPP)
responded positively to availabile soil moisture, but the response did
not vary across treatments. There was also no evidence that treatment
effects grew over time. The similarity of ecosystem functional response
across treatment was not due to compensatory shifts at the plant
community level, where species composition among treatments was similar
and remarkably stable over the five years. At least in the short-term,
ecosystem functional response and community composition in this
sagebrush steppe system is resistant to increases and decreases in
growing season precipitation.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

To that end, here we report the results of a five-year precipitation
manipulation experiment in a sagebrush steppe grassland. We imposed
drought and irrigation treatments (50\% decrease/increase) and measured
ecosystem (ANPP) and community (species composition) responses. We focus
on how the drought and irrigation treatments affect the relationship
between available soil moisture and ANPP, and if community dynamics
underly the ecosystem responses. In particular, we are interested in the
consistency of the soil moisture-ANPP relationship among treatments. Is
the relationship steeper under the drought treatment, at low soil
moisture? Does the relationship saturate under the irrigation treatment,
at high soil moisture? To answer these questions we fit a random
intercepts, random slopes model to test whether the regressions differed
among treatments. We also analyzed community composition over time,
allowing us to place our experimental results within the framework of
the HRF and our competing predictions (Fig. 1).

@@

We estimated aboveground net primary productivity (ANPP) using a
radiometer to relate ground reflectance to plant biomass \citep[see][
for a review]{Byrne2011}. We recorded ground reflectance at four
wavelengths, two associated with red reflectance (626 nm and 652 nm) and
two associated with near-infrared reflectance (875 nm and 859 nm). At
each plot in each year, we took four readings of ground reflectances at
the above wavelengths. We also took readings in ten calibration plots
adjacent to the experimental site, in which we harvested all aboveground
biomass, dried it to a constant weight at 60\(^{\circ}\)C, and weighed
it to estimate ANPP.

For each plot and year, we averaged the four readings for each
wavelength and then calculated NDVI using the MODIS and AVHRR
algorithms. To convert NDVI to ANPP we regressed NDVI against the dry
biomass weight from the ten calibration plots. We fit regressions to
MODIS-based NDVI and AVHRR-based NDVI for each year and retained the
regression with the better fit. Using the best regression equation for
each year, we predicted ANPP (Appendix 1).

Species composition data came from two sources: yearly census maps for
each plot made using a pantograph \citep{Hill1920} and yearly counts of
annual species in each plot. The maps record the spatial location of all
individuals in the plot and the basal cover of each individual with
cover greater than 1 cm. Using those annual maps, we aggregated over
individuals to calculate total basal cover for each species in each
plot. We made a large plot-treatment-year by species matrix, where
columns were filled with either basal cover or density, depending on the
measurement made for the particular species. So we could analyze the
different types of data together, we standardized the values in each
column. This puts all abundance values on the same scale, but comes with
the limitation that all species are weighted equally. Nonetheless, this
scaling approach allows a comprehensive view of community composition
dynamics through time.

@@

Our main goal was to test whether the relationship between ANPP and
growing season precipitation (hereafter, precipiation) differed among
the drought, control, and irrigation treatments. To achieve this goal,
we fit a multi-level random intercept and random slope regression with
log(ANPP) as the response variabile and soil moisture (volumetric water
content) as the sole predictor. We fit the model under a Bayesian
framework, allowing us to test for treatment differences by comparing
the posterior distributions of the treatment-level coefficients
\citep[e.g.,][]{Tredennick2013}. Both log(ANPP) and soil moisture were
standardized to have mean 0 and unit variance before fitting the model
{[}i.e., \((x_i - \bar{x})/\sigma_x\){]}.

Our multi-level model has three grouping levels for coefficients,
representing the nested structure of the data: (i) overall coefficients,
(ii) treatment coefficients, and (iii) plot coefficients. Each
subsequent level is drawn from the distribution of coefficients at the
previous level. Formally, our model is defined as follows:

@@

\begin{align}
\mu_{i(j(k(t)))} &= \beta_{0,j(k)} + \beta_{1,j(k)}x_i + \gamma_t, \\
y_{i(j(k(t)))} &\sim \text{Normal} \left(\mu_{i(j(k(t)))}, \sigma^2_{k} \right),
\end{align}

\noindent{}where \(\mu_{i(j(k))}\) is the deterministic prediction from
the regression model for observation \emph{i} for plot \emph{j}
associated with treatment \emph{k} in year \emph{t}, \(\beta_{0,j(k)}\)
is the intercept for plot \emph{j} associated with treatment \emph{k},
\(\beta_{1,j(k)}\) is the slope term for the effect of soil moisture for
plot \emph{j} associated with treatment \emph{k}, \(\gamma_t\) is the
intercept offset for year \emph{t}, and \(\sigma^2_k\) is the process
variance for treatment \emph{k}. Data include the standardized log(ANPP)
observations \(\left( y_{i(j(k(t)))} \right)\) and soil moisture
(\(x_i\)). Although we include observation subscript \emph{i} on the
\emph{x}s, observations within a treatment-year all share the same soil
moisture values.

The intercept and slope terms are modeled hierarchichally to account for
the non-independence of observations across years within plots and to
allow us to test the hypothesis that our treatments alter the ANPP-soil
moisture relationship. As noted above, plot-level coefficients are drawn
from treatment-level coefficients, which are drawn from overall
coefficients. We also include a covariance structure among the intercept
and slope at each level. Formally, our hierarhical structure is as
follows, where we drop the intercept (0) and slope (1) subscripts and
instead refer to a vector of coefficients, \(\boldsymbol{\beta}\):

@@

Cumulative growing season soil moisture had a positive effect on ANPP
(mean of \(\beta_{1}\) = 0.41; 80\% BCI = -0.08, 0.86; 95\% BCI = -0.45,
1.13) (Fig. 3). Ecosystem functional response was similar among
treatments, with treatment level regressions between soil moisture and
log(ANPP) having similar intercepts and slopes (Fig. 3A). There was also
no evidence that the treatment effects became more important over time
because there was no directional trend in the random year effects (Fig.
4).

@@

Third, the ecosystem and community resistance to drought and deluge we
observed could be a real phenomenon in this system, at least over the
five year time span we observed. The perennial plants in this cold
desert ecosystem are tolerant to drought conditions \citep[A.R.
Kleinhesselink, unpublished data]{Bazzaz1979, Franks2011}, meaning a
consistent ecosystem functional response at low soil moisture is
perfectly reasonable. For example, many of the perennial grasses in our
focal ecosystem avoid drought stress by growing early in the growing
season (A.R. Kleinhesselink, personal observation). Likewise, the
dominant shrub in our focal ecosystem, \emph{Artemisia tripartita}, has
access to water deep in the soil profile thanks to a deep root system
\citep{Germino2014}. As for the irrigation treatments, the lack of a
saturating response (e.g., a lower regression slope) suggests water
availability remained a limiting factor at the levels of precipitation
we were able to manipulate.

>>>>

\usepackage{setspace} \usepackage{todonotes} \usepackage{rotating}
\usepackage{color, soul} \usepackage{sectsty} \usepackage{bm,mathrsfs}

@@

% 
\begin{abstract}
Precipitation is predicted to become more variable in the western United
States, meaning years of above and below average precipitation will
become more common. Periods of drought and deluge are major drivers of
ecosystem functioning in water limited grasslands, but how ecosystems
respond to precipitation may change as the duration of above and below
average periods increases. Changes in ecosystem functional response
could reflect underlying changes in species composition or species
reaching physiological thresholds at extreme precipitation levels. Here
we report the results of a five-year experiment where we used drought
and irrigation treatments (approximately 50\% decrease/increase) to
investigate whether ecosystem functional response remains consistent at
extreme precipitation levels. We also use species-level adundance data
to link ecosystem functional response to community dynamics. Ecosystem
functional response, defined as the relationship between soil moisture
and aboveground net primary productivity, was suprisingly consistent
across drought, control, and irrigation treatments. Aboveground net
primary productivity (ANPP) responded positively to availabile soil
moisture, but the response was weak and did not vary across treatments.
There was also no evidence that treatment effects grew over time. The
similarity of ecosystem functional response across treatments was not
due to compensatory shifts at the plant community level, where species
composition among treatments was similar and remarkably stable over the
five years. At least in the short-term, ecosystem functional response
and community composition in this sagebrush steppe system is resistant
to increases and decreases in growing season precipitation, perhaps due
to bet-hedging strategies of constituent species.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

To that end, here we report the results of a five-year precipitation
manipulation experiment in a sagebrush steppe grassland. We imposed
drought and irrigation treatments (approximately \(\pm50\%\)) and
measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the
relationship between available soil moisture and ANPP, and if community
dynamics underly the ecosystem responses. In particular, we are
interested in the consistency of the soil moisture-ANPP relationship
among treatments. Is the relationship steeper under the drought
treatment, at low soil moisture? Does the relationship saturate under
the irrigation treatment, at high soil moisture? To answer these
questions we fit a random intercepts, random slopes model to test
whether the regressions differed among treatments. We also analyzed
community composition over time, allowing us to place our experimental
results within the framework of the HRF and our competing predictions
(Fig. 1).

@@

We estimated aboveground net primary productivity (ANPP) using a
radiometer to relate ground reflectance to plant biomass \citep[see][
for a review]{Byrne2011}. We recorded ground reflectance at four
wavelengths, two associated with red reflectance (626 nm and 652 nm) and
two associated with near-infrared reflectance (875 nm and 859 nm). At
each plot in each year, we took four readings of ground reflectances at
the above wavelengths. We also took readings in 12 (2015), 15 (2012,
2013, 2014), or 16 (2016) calibration plots adjacent to the experimental
site, in which we harvested all aboveground biomass, dried it to a
constant weight at 60\(^{\circ}\)C, and weighed it to estimate ANPP.

For each plot and year, we averaged the four readings for each
wavelength and then calculated a greenness index based on the same bands
used to calculate NDVI using the MODIS and AVHRR bands for NDVI. To
convert the greenness index to ANPP we regressed the greenness index
against the dry biomass weight from the ten calibration plots. We fit
regressions to MODIS-based index and AVHRR-based index for each year and
retained the regression with the better fit. Using the best regression
equation for each year, we predicted ANPP (Appendix 1).

Species composition data came from two sources: yearly census maps for
each plot made using a pantograph \citep{Hill1920} and yearly counts of
annual species in each plot. From those maps, we have data on the
density of all annuals and perennials forbs, basal cover of perennial
grasses, and canopy cover of shrubs. We made a large plot-treatment-year
by species matrix, where columns were filled with either basal cover or
density, depending on the measurement made for the particular species.
So we could analyze the different types of data together, we
standardized the values in each column. This puts all abundance values
on the same scale, meaning that common and rare species are weighted
equally. Nonetheless, if we assume that rare species will respond more
than common ones, then this approach is anti-conservative for detecting
community change. This means that our approach is biased toward
detecting compositional changes.\todo{Did I say this right?}

@@

Our main goal was to test whether the relationship between ANPP and soil
moisture differed among the drought, control, and irrigation treatments.
Based on our own observations and previous work at our study site
\citep{Blaisdell1958, Dalgleish2011, Adler2012}, we chose to use
cumulative volumetric water content from March through June as our
metric of soil moisture (hereafter referred to simply as `soil
moisture'). To achieve this goal, we fit a multi-level random intercept
and random slope regression with log(ANPP) as the response variabile and
soil moisture as the sole predictor. We log-transformed ANPP to account
for heteroscedasticity. We fit the model under a Bayesian framework,
allowing us to test for treatment differences by comparing the posterior
distributions of the treatment-level coefficients
\citep[e.g.,][]{Tredennick2013}. Both log(ANPP) and soil moisture were
standardized to have mean 0 and unit variance before fitting the model
{[}i.e., \((x_i - \bar{x})/\sigma_x\){]}.

Our model is defined as follows:

@@

\begin{align}
\mu_{j(k(t))} &= \beta_{0,j(k)} + \beta_{1,j(k)}x_{k(t)} + \gamma_t, \\
y_{j(k(t))} &\sim \text{Normal} \left(\mu_{j(k(t))}, \sigma^2_{k} \right),
\end{align}

\noindent{}where \(\mu_{j(k(t))}\) is the deterministic prediction from
the regression model for plot \emph{j} associated with treatment
\emph{k} in year \emph{t}, \(\beta_{0,j(k)}\) is the intercept for plot
\emph{j} associated with treatment \emph{k}, \(\beta_{1,j(k)}\) is the
slope term for the effect of soil moisture for plot \emph{j} associated
with treatment \emph{k}, \(\gamma_t\) is the intercept offset for year
\emph{t}, and \(\sigma^2_k\) is the process variance for treatment
\emph{k}. Data include the standardized log(ANPP) observations
\(\left( y_{j(k(t))} \right)\) and soil moisture (\(x_{k(t)}\)).

The intercept and slope terms are modeled hierarchichally to account for
the non-independence of observations across years within plots and to
allow us to test the hypothesis that our treatments alter the ANPP-soil
moisture relationship. Thus, our multi-level model has three grouping
levels for coefficients, representing the nested structure of the data:
(i) overall coefficients, (ii) treatment coefficients, and (iii) plot
coefficients. Each subsequent level is drawn from the distribution of
coefficients at the previous level, meaning plot-level coefficients are
drawn from treatment-level coefficients, which are drawn from overall
coefficients. We also include a covariance structure among the intercept
and slope at each level. Formally, our hierarhical structure is as
follows, where we drop the intercept (0) and slope (1) subscripts and
instead refer to a vector of coefficients, \(\boldsymbol{\beta}\):

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP
(mean of \(\beta_{1}\) = 0.41), but the effect is associated with high
uncertainty, with credible intervals that overlap zerp (80\% BCI =
-0.08, 0.86; 95\% BCI = -0.45, 1.13) (Fig. 3). Ecosystem functional
response was similar among treatments, with treatment level regressions
between soil moisture and log(ANPP) having similar intercepts and slopes
(Fig. 3A). There was also no evidence that the treatment effects became
more important over time because there was no directional trend in the
random year effects (Fig. 4).

@@

Third, the ecosystem and community resistance to drought and deluge we
observed could be a real phenomenon in this system, at least over the
five year time span we observed. The perennial plants in this cold
desert ecosystem are tolerant to drought conditions \citep[A.R.
Kleinhesselink, unpublished data]{Bazzaz1979, Franks2011}, meaning a
consistent ecosystem functional response at low soil moisture is
perfectly reasonable. For example, many of the perennial grasses in our
focal ecosystem avoid drought stress by growing early in the growing
season \citep[A.R. Kleinhesselink, personal observation]{Blaisdell1958}.
Likewise, the dominant shrub in our focal ecosystem, \emph{Artemisia
tripartita}, has access to water deep in the soil profile thanks to a
deep root system \citep{Germino2014}. As for the irrigation treatments,
the lack of a saturating response (e.g., a lower regression slope)
suggests water availability remained a limiting factor at the levels of
precipitation we were able to manipulate.
